Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
revision to conclusion

\section{Future work and Conclusions}
\label{sec:conclusion}
In this work, we present a short and mechanized proof of consistency through
proof by logical relations for a dependently typed language with a full
universe hierarchy, large eliminations, an intensional identity type, and
dependent eliminators.  We show the extensibility of our approach by proving
the existence of $\beta\eta$-normal forms with only small and mechanical
changes to our proof development. Related work gives us confidence that we
could extend our logical relation to include features such as full inductive
datatypes, irrelevant arguments, and type-directed conversion; though it is
not clear how much of the brevity of the development we will be able to
maintain.  Furthermore, we hope that mechanized logical relations proofs will
eventually grow to include other features found in dependent type theories,
such as impredicative universes, universe polymorphism, and cumulativity.

Our Coq mechanization leverages existing Coq libraries for reasoning about
metatheory and for general purpose automation, allowing us to significantly
reduce the verbosity typically associated with mechanized proofs. The result
is a declarative proof style that is usually only available on pen and
paper. Our development shows that proofs by logical relations for dependent
types are accessible and do not require months of effort to implement.
We hope our proof can inspire researchers to more frequently mechanize
results, such as consistency and normalization, for their dependent type
theories. By design, the goals of our work are modest.

>>>>

\section{Future work and Conclusions}
\label{sec:conclusion}
In this work, we present a short and mechanized proof of consistency through
proof by logical relations for a dependently typed language with a full
universe hierarchy, large eliminations, an intensional identity type, and
dependent eliminators.  We show the extensibility of our approach by proving
the existence of $\beta\eta$-normal forms with only small and mechanical
changes to our proof development.  Our Coq mechanization leverages existing
Coq libraries for reasoning about metatheory and for general purpose
automation, allowing us to significantly reduce the verbosity typically
associated with mechanized proofs. The result is a declarative proof style
that is usually only available on pen and paper.

Related work gives us confidence that we could extend our logical relation to
include features such as full inductive datatypes, irrelevant arguments, and
type-directed conversion; however, it is not clear how much of the brevity of
this development can be maintained.  Furthermore, we hope that mechanized
logical relations proofs will eventually grow to include other features found
in dependent type theories, such as impredicative universes, universe
polymorphism, and cumulativity.  Regardless, our development shows that proofs
by logical relations for dependent types are accessible and do not require
months of effort to implement.  We hope our proof can inspire researchers to
more frequently mechanize results, such as consistency and normalization, for
their dependent type theories.
